---
title: "Class Notes 3.28.35 - 04.02.25"
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(readxl)
library(modeldata)

pokemon <- read_csv("pokemon (1).csv")
trout <- read_excel("trout.xlsx")
```

Merges: combine two files that you have been working on

## Git

We talked about branches a bit. See my video on git branches and merges. 

## Anova

When to use Anova:  
Inference with 2 variables
1. 2 categorical (x^2)  
2. 2 quantitative (regression)  
3. 1 categorical and 1 quantitative (anova)  

We'll use the `pokemon` data set from Moodle. Also see my slide on anova on Moodle.

Is `hp` consistent, on average, across generations?

Null: the mean hp is the same for each generation (independent).

Alternative: the mean hp is different for at least one generation (dependence).

```{r}
ggplot(pokemon,
       aes(x = hp,
           y = as.factor(generation))) + #as.character() also works here
  geom_boxplot()
```

To run the test, we need `generation` to be categorical, so we do a `mutate` before the anova.

```{r}
pokemon <- pokemon %>% 
  mutate(generation = as.factor(generation))

aov(pokemon$hp ~ pokemon$generation)
```

To see the p-value, we save this output and then get a `summary`.

```{r}
model <- aov(hp ~ generation, 
             data = pokemon) # same aov
summary(model)
```

The p-value (.088) is not very small. There is only very weak evidence that the variables are dependent on one another. It appears that `generation` does not inform `hp`.

## Example 2: trout

```{r}
glimpse(trout)
```

Are these variables independent?

$H_0:$ `Hemo` is the same at different locations, on average.

$H_a:$ At least one group mean is different.

```{r}
trout <- trout %>% 
  mutate(Location = as.factor(Location))

ggplot(trout,
       aes(x = Location,
           y = Hemo)) + 
  geom_boxplot() + 
  geom_jitter()
```

Let's run the anova.

```{r}
model <- aov(Hemo ~ Location,
             data = trout)
summary(model)
```

The p-value is very small, providing good evidence against the null. We can conclude that Hemo is informed by Location.

(Last thing today: commit changes to the git repo) 

## Apr. 2: Post-hoc testing

If the anova leads to rejecting the null, we can consider pair-wise comparisons of the groups. HOWEVER, we have to be honest about the fact that there are many possible comparisons. One very common way to do this is the **Tukey Honest Standard Differences Test**.

```{r}
TukeyHSD(model)
```

## Comparing two quantitative variables

```{r}
glimpse(crickets)
```

Null hypothesis: `rate` is independent of `temp`.

Alternative: `rate` is not independent of `temp`.

More specifically, we'll test for a linear dependence. That is, we'll test if the slope of regression line is nonzero.

```{r}
ggplot(crickets, 
       aes(x = temp,
           y = rate)) + 
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE)
```

More specific hypothesis:

Null: the slope is zero in the population from which we are sampling.

Alternative: the slope is nonzero

Let's run this test.

```{r}
lm(rate ~ temp,
   data = crickets) # syntax just like aov
```

The equation of the regression line rate = -27.948 + 4.243*temp. We extract more information (including a p-value for our test) by saving the model and getting a summary.

```{r}
model <- lm(rate ~ temp,
   data = crickets)
summary(model)
```

The p-value for the slope is infinitesimal. We reject the null and conclude there is a non-zero slope, that rate is dependent on temp.






